From Net-Zero to Net-Negative: Assessing \hydrogen{} and \carbon{} Network Strategies in Europe

### 2023-11-26 report/submissions/journal/paper/paper.tex
incorporate feedback from FN in Results

To supply carbon needed for CS and CU, the system can choose to deploy carbon capture technology at various point sources (see below), or through DAC facilities. The concept of a ``merit order for captured carbon`` plays a pivotal role in optimizing and prioritizing the deployment of carbon capture technologies based on their economic feasibility. This merit order ranks all CC technologies according to their relative costs to capture a ton of carbon (see Fig.~X). At the lower end of the cost spectrum, CC technologies applied to process emissions, such as those from cement, offer a cost-effective starting point. Following this, biomass combined heat and power (CHP) systems provide a dual benefit of energy production and carbon capture. Moving up the scale, gas used in industrial applications and biomass employed for industrial processes represent more costly yet viable options to capture carbon. DAC, an emerging technology capable of extracting \carbon{} directly from the atmosphere, stands at a higher cost level due to its current high capital costs and energy demand. Finally, based on our model results, biogas-to-gas upgrading, a process that refines biogas to natural gas quality, incurs the highest costs in the merit order per ton of captured \carbon{}. Biogas input is a high-cost fuel which based on the endogenous modeling decisions is not used in large quantities. Thus, the marginal price of capturing additional \carbon{} from biogas-to-gas upgrading is high because a substantial amount of the biogas fuel costs factors into the cost of captured \carbon{}. This merit order framework is crucial for strategically deploying necessary carbon capture solutions while balancing economic considerations. If we consider the spatial aspect of distributed carbon capture potentials, the \carbon{} network can exploit comparative cost advantages between ``CO2 bidding zones'' to utilize the lowest-cost CC facilities across the continent. We assume a capture rate of 90\% for CC on process emissions, SMR, biogas-to-gas, as well as gas and biomass used in industry, and 95\% for CC on biomass and gas CHPs.

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/cost_bar.png}
    \caption[short]{Total annual system cost for the European energy system for the different models subdivided into groups of technologies. While the `Baseline' model has neither a \carbon{} nor a \hydrogen{} network, `Hybrid' is allowed to expand both. "Gas Infrastructure" combines gas facilities for power and heat production,"\carbon{} Infrastrucure" combines transport, storage and sequestration, "H$_2$ Infrastructure" combines transport and storage. "Carbon Capture at Point Sources" combines all technologies with integrated carbon capture.}
    \label{fig:cost_bar}
\end{figure}

System cost differences between the models are driven by renewable energy resources (solar, wind, and bioenergy), the \carbon{} and \hydrogen{} networks, as well as carbon capture technologies (DAC and CC at point sources).
In the Baseline model, the absence of dedicated carbon and hydrogen transport networks requires the system to deploy carbon production technologies (DAC and CC at point sources) in the regions with carbon sequestration sites, and small amounts of hydrogen production technologies (mainly electrolysis) in each region to meet local hydrogen demand for industry and land transport fuel cells.
For CU and its respective carbon and hydrogen demands, the system chooses the sites where on-site production of carbon, hydrogen and the synthesis process has the lowest cost. This leads to renewable resources and storage solutions being deployed not necessarily in the regions with the best resources or lowest costs for clean electricity generation, but where they are situated closer to the locations with carbon supply from CC at point sources, and to the locations of energy demand for electrolysers and DAC.
Thus, total system costs in the Baseline model are highest, while the models with additional transport options for carbon and hydrogen achieve a more efficient system allocation and lower total system costs.

We can now closer examine the cost reductions and shifts in cost allocation when shifting from the Baseline model to the other three network models.
The implementation of a \carbon{} transport network (\textit{\carbon{}~Grid}) facilitates efficient point-source carbon capture, substantially reducing the dependency on DAC technologies. While DAC supplies around 50\% of the carbon needed in the Baseline model (400 Mt \carbon{}, see Fig.~\ref{fig:balance_captured_carbon}), in the \modCO{}, DAC is only used to supply around 20\% of the carbon (150 Mt \carbon{}). The remaining carbon is supplied by additional carbon capture technology on all plants burning biomass in industry facilities and in CHPs as well as all industry facilities using gas (see Fig.~\ref{fig:captureshare_line}), because the cost to deploy on-site carbon capture technology has relatively lower costs than DAC. The \carbon{} transport network enables the transport of carbon from the locations of the lowest-cost captured carbon sources to the locations of the carbon sinks at CU and CS sites, reducing the need for higher cost DAC. The improved carbon management infrastructure also allows for a more strategic deployment of renewables, particularly in areas that combine high quality renewables and high sequestration potentials, leading to system-wide cost savings in renewable generation and storage.

@@

The \hydrogen{}-Grid model underscores the pivotal role of hydrogen infrastructure. The creation of a hydrogen transport network enables the cost-effective delivery of hydrogen from regions with a surplus of low-cost renewable energy to high-demand industrial centers. This model shows a pronounced decrease in investments in solar generation and gas infrastructure. The \hydrogen{} network reduces DAC similarly to the \carbon{} network, but still produces 200 Mt of carbon from DAC. CC technology on process emissions increases from 90\% to 100\%, on biomass CHPs from 60\% to 100\%, and on facilities using gas in industry from 50\% to 75\%. CC on biomass use in industry remains at 50\%, same as the \modCO{} (see Fig.~\ref{fig:captureshare_line}). This variation highlights how the implementation of \carbon{} and \hydrogen{} networks unlocks specific carbon capture potentials.

While the cost reduction in the Hybrid model compared to the \hydrogen-Grid model is marginal, the allocation of costs reveals a propagation of reduced DAC facilities and increased CC at point sources. The Hybrid model relies slightly less on investments into the \hydrogen{} network than the \modH, but adds some investments into the \carbon{} network. Still, investments into the \hydrogen{} network outweigh investments into the \carbon{} network by a factor of four.

>>>>

To supply carbon needed for CS and CU, the system can choose to deploy carbon capture technology at various point sources (see below), or through DAC facilities. The concept of a ``merit order for captured carbon`` plays a pivotal role in optimizing and prioritizing the deployment of carbon capture technologies based on their economic feasibility. This merit order ranks all CC technologies according to their relative costs to capture a ton of carbon (see Fig.~X). At the lower end of the cost spectrum, CC technologies applied to process emissions, such as those from cement, offer a cost-effective starting point. Following this, biomass combined heat and power (CHP) systems provide a dual benefit of energy production and carbon capture. Moving up the scale, gas used in industrial applications and biomass employed for industrial processes represent more costly yet viable options to capture carbon. DAC, an emerging technology capable of extracting \carbon{} directly from the atmosphere, stands at a higher cost level due to its current high capital costs and energy demand. Finally, based on our model results, biogas-to-gas upgrading, a process that refines biogas to natural gas quality, incurs the highest costs in the merit order per marginal ton of captured \carbon{}. Biogas input is a high-cost fuel which based on the endogenous modeling decisions is not used in large quantities. To the extent that biogas-to-gas facilities are used by the model to supply additional (carbon-neutral) gas as fuel, adding CC infrastructure incurs low costs (and thus ranges on the left-hand side of the merit order curve). However, the price of capturing additional \carbon{} from biogas-to-gas upgrading is high because a substantial amount of the biogas fuel costs factors into the marginal cost of captured \carbon{}. This merit order framework is crucial for strategically deploying necessary carbon capture solutions while balancing economic considerations. If we consider the spatial aspect of distributed carbon capture potentials, the \carbon{} network can exploit comparative cost advantages between ``CO2 bidding zones'' to utilize the lowest-cost CC facilities across the continent. We assume a capture rate of 90\% for CC on process emissions, SMR, biogas-to-gas, as well as gas and biomass used in industry, and 95\% for CC on biomass and gas CHPs.

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/cost_bar.png}
    \caption[short]{Total annual system cost for the European energy system for the different models subdivided into groups of technologies. While the Baseline model has neither a \carbon{} nor a \hydrogen{} network, the Hybrid model is allowed to expand both. ``Gas Infrastructure'' combines gas facilities for power and heat production, ``\carbon{} Infrastrucure'' combines transport, storage and sequestration, and ``H$_2$ Infrastructure'' combines transport and storage. ``Carbon Capture at Point Sources'' combines all technologies with integrated carbon capture, including the cost of the main facility (e.g., CHP unit) and the carbon capture application.}
    \label{fig:cost_bar}
\end{figure}

System cost differences between the models are driven by renewable energy resources (solar, wind, and bioenergy), the \carbon{} and \hydrogen{} networks, as well as carbon capture technologies (DAC and CC at point sources).
In the Baseline model, the absence of dedicated carbon and hydrogen transport networks requires the system to deploy carbon production technologies (DAC and CC at point sources) in the regions with carbon sequestration sites, and distributed amounts of hydrogen production technologies (mainly electrolysis) in each region to meet local hydrogen demand for industry and land transport fuel cells.
For CU and its respective carbon and hydrogen demands, the system chooses the sites where on-site production of carbon, hydrogen as well as the inputs and by-product outputs (e.g., waste heat) of the synthesis process have the lowest cost (or highest value). This leads to renewable resources and storage solutions being deployed not necessarily in the regions with the best resources or lowest costs for clean electricity generation, but where they are situated closer to the endogenously defined locations of energy demand for electrolysers and DAC.
Thus, total annual system costs in the Baseline model are highest, totalling \label{}803 billion euros (see Fig.~\ref{fig:objective_heatmap}). The models with additional transport options for carbon and hydrogen achieve a more efficient system allocation and lower total system costs, with around \label{}2\% cost reductions (\label{}781 billion euros) for the \carbon{} network, and around \label{}4\% cost reductions for the \hydrogen{} network and the hybrid scenario (\label{}765 and \label{}764 billion euros, respectively).

We can now closer examine the cost reductions and shifts in cost allocation when shifting from the Baseline model to the other three network models.
The implementation of a \carbon{} transport network (\textit{\carbon{}~Grid}) facilitates carbon capture from the point sources with the lowest costs across regions, substantially reducing the dependency on DAC. While DAC supplies around 50\% of the carbon needed in the Baseline model (400 Mt \carbon{}, see Fig.~\ref{fig:balance_captured_carbon}), in the \modCO{}, DAC is only used to supply around 20\% of the carbon handled by the system (150 Mt \carbon{}). The remaining carbon is supplied by additional carbon capture technology on all plants burning biomass in industry facilities and in CHPs as well as all industry facilities using gas (see Fig.~\ref{fig:captureshare_line}), because the cost to deploy on-site carbon capture technology has relatively lower costs than DAC. The \carbon{} transport network enables the transport of carbon from the locations of the lowest-cost captured carbon sources to the locations of the carbon sinks at CU and CS sites, reducing the need for higher cost DAC. The option to transport carbon also allows for the deployment of renewables in areas that combine high quality renewable resources and high sequestration potentials, leading to system-wide cost savings in renewable generation and storage.

@@

The \hydrogen{}-Grid model underscores the beneficial role of hydrogen infrastructure. The creation of a hydrogen transport network enables the cost-effective delivery of hydrogen from regions with a surplus of low-cost renewable energy to high-demand industrial centers. This model shows a pronounced decrease in investments in DAC, and accordingly also solar generation which provides most of the electricity to run DAC facilities. The reduction in DAC is similar to the \carbon{} network, but the \hydrogen{} network still produces 200 Mt of carbon from DAC. CC technology on process emissions increases from 90\% to 100\%, on biomass CHPs from 60\% to 100\%, and on facilities using gas in industry from 50\% to 75\%. CC on biomass use in industry remains at 50\%, same as the \modCO{} (see Fig.~\ref{fig:captureshare_line}). This is because the \hydrogen{} network can supply low-cost \hydrogen{} to \carbon{} point sources, where without the \hydrogen{} network, fuel synthethis with locally captured \carbon{} and locally sourced electrolytic hydrogen would have been more expensive than synthesizing fuel with \carbon{} from DAC and hydrogen from electrolysis at the best renewable resource sites. This highlights how the implementation of \carbon{} and \hydrogen{} networks unlocks carbon capture potentials at distributed point sources across the continent.

While the cost reduction in the Hybrid model compared to the \hydrogen-Grid model is marginal, the allocation of costs reveals a propagation of reduced DAC facilities and increased CC at point sources. In other words, the \carbon{} network allows to tap into far away biomass potentials to capture carbon indirectly from the air, which provides marginally lower costs than DAC, even considering carbon transport costs. The Hybrid model relies slightly less on investments into the \hydrogen{} network than the \modH, but adds some investments into the \carbon{} network. Still, investments into the \hydrogen{} network outweigh investments into the \carbon{} network by a factor of four.
